---
author: Jose Toledo Luna
date: November 09, 2023
title: One Sample $t$-test
toc: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  eval = TRUE
)
```

<details>
<summary>Required Packages</summary>
```{r,eval=FALSE}
library(openintro)
```
</details>

---

## Hypothesis tests

There are a few components common to every hypothesis test

**Null Hypothesis**

A hypothesis that posits there is no significant difference, effect, or relationship in the variables being studied. Suggesting any observed differences are due to random chance Denoted by $H_0$ ("H naught")


**Alternative Hypothesis**

A hypothesis that contradicts the null hypothesis, proposing that there is a significant difference, effect, or relationship in the variables being studied, beyond what could be explained by random chance. Denoted by $H_a$

**Test Statistic**

A numerical value calculated from sample data used to assess the strength of the evidence against the null hypothesis, with larger values indicating stronger evidence in favor of the alternative hypothesis

**p-value**

The probability of a test statistic as rare or even more rare than the one observed under the assumptions of the null hypothesis


The p-value quantifies the likelihood of observing the data or more extreme results under the null hypothesis; a lower p-value indicates that the observed statistic is highly improbable under the null hypothesis, strengthening the evidence against it and often leading to its rejection in favor of the alternative hypothesis



## One Sample $t$-test

To perform a theory-based hypothesis test for the mean of a quantitative variable, following the components of a hypothesis test, but in the context of a one sample $t$-test


1. State the null and alternative hypotheses about the mean parameter $\mu$

2. From a sample of data, compute the sample mean $\bar{x}$ and its standardized $t$-statistic

3. Check the validity conditions to apply the Central Limit Theorem

4. If the validity conditions hold, compute the $p$-value by comparing the observed $t$-statistic to a $t$-distribution with $df = n-1$


---

We can use the `t.test()` function to conduct this hypothesis test

The basic syntax for the `t.test()` function is as follows:

```{r,eval=FALSE}
t.test(x, y = NULL, 
       alternative = c("two.sided", "less", "greater"),
       mu = 0, conf.level = 0.95, ...)
```

* `x`: numeric vector of data values *(our sample)*
* `y`: an *optional* numeric vector of data values *(response)*
* `alternative`: type of hypothesis being performed, 
  - `"two.sided"`: $H_a: \mu \neq \mu_0$
  - `"less"`: $H_a: \mu < \mu_0$
  - `"greater"`: $H_a: \mu > \mu_0$
* `mu`: population mean $\mu$ *(null value)*, can be difference of means if performing a two-sample test
* `conf.level`: confidence level of the interval $100(1-\alpha)$%, where $\alpha$ is the significance level

For additional arguments `...` see `?t.test()`


---


Consider the `Tip Data` from `opentintro` package. The tip data is a simulated data set
of tips over a few weeks on a couple days per week. Each tip is associated with a single group, which may include several bills and tables


From this dataset we obtain a simple random sample of 30 tips, but first we only consider tips from bills that were at most $80

```{r}
tips <- openintro::tips
```

```{r,echo=FALSE}
dplyr::glimpse(tips)
```


```{r}
tips_bill_at_most80 <- tips[tips$bill <= 80,]
```


Among the tips from bills that were at most \$80, we obtained a simple random sample of size 30. We are interested in determining if the average tips throughout the weeks will be greater than \$5 per group *(for some reason, the business needs on average at least \$5 tips per group to stay in business)*. 

The null and alternative hypothesis are given by
$$
\begin{align*}
H_0: \mu = 5 \\[5pt]
H_a: \mu > 5
\end{align*}
$$



```{r}
set.seed(1234)
rows_sample <- sample(1:nrow(tips_bill_at_most80), 30)
tips_sample <- tips_bill_at_most80[rows_sample,]
```

First we will carry out a one-sample $t$-test the long way. First, we compute the sample mean for the tips

```{r}
x_bar <- mean(tips_sample$tip)
x_bar
```

Next, compute the sample standard deviation of the tips


```{r}
s <- sd(tips_sample$tip)
```

the $t$-statistic is then computed as 

$$
t = \frac{\bar{x}-\mu_0}{\left(\frac{s}{\sqrt{n}} \right)}
$$

```{r}
n <- length(rows_sample)
mu_0 <- 5
t_stat <- (x_bar - mu_0) / (s / sqrt(n))
```


```{r}
t_stat
```

Lastly, obtain the $p$-value

```{r}
p_value <- pt(t_stat, df = n - 1,lower.tail = FALSE)
```

```{r}
p_value
```


At a significance level of $\alpha=0.05$, we reject the null hypothesis, indicating there is enough evidence to propose that, on average, the tips per group exceed \$5 for bills equal to or less than \$80, as supported by the three-week data provided


With a single command, we can use `t.test()` function and obtain all of the following information, without having to compute every step ourselves

- observed sample mean $\bar{x}$
- $t$-statistic
- degrees of freedom $(df)$
- the alternative hypothesis we are testing
- $p$-value
- A 95% confidence interval for $\mu$

```{r}
t.test(tips_sample$tip, mu=5, 
       alternative = 'greater')
```


