---
author: Jose Toledo Luna
date: November 09, 2023
title: One Sample $t$-test
toc: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  eval = TRUE
)
```

<details>
<summary>Required Packages</summary>
```{r}
library(ggplot2)
library(patchwork)
```


```{r}
theme_set(theme_bw())
theme_replace(panel.grid.minor = element_blank(),
              panel.grid.major = element_blank())
```
</details>

---

## Hypothesis tests

There are a few components common to every hypothesis test

**Null Hypothesis**

A hypothesis that posits there is no significant difference, effect, or relationship in the variables being studied. Suggesting any observed differences are due to random chance Denoted by $H_0$ ("H naught")


**Alternative Hypothesis**

A hypothesis that contradicts the null hypothesis, proposing that there is a significant difference, effect, or relationship in the variables being studied, beyond what could be explained by random chance. Denoted by $H_a$

**Test Statistic**

A numerical value calculated from sample data used to assess the strength of the evidence against the null hypothesis, with larger values indicating stronger evidence in favor of the alternative hypothesis

**p-value**

The probability of a test statistic as rare or even more rare than the one observed under the assumptions of the null hypothesis


The p-value quantifies the likelihood of observing the data or more extreme results under the null hypothesis; a lower p-value indicates that the observed statistic is highly improbable under the null hypothesis, strengthening the evidence against it and often leading to its rejection in favor of the alternative hypothesis



## One Sample $t$-test

To perform a theory-based hypothesis test for the mean of a quantitative variable, following the components of a hypothesis test, but in the context of a one sample $t$-test


1. State the null and alternative hypotheses about the mean parameter $\mu$

2. From a sample of data, compute the sample mean $\bar{x}$ and its standardized $t$-statistic

3. Check the validity conditions to apply the Central Limit Theorem

4. If the validity conditions hold, compute the $p$-value by comparing the observed $t$-statistic to a $t$-distribution with $df = n-1$



---

We can use the `t.test()` function to conduct this hypothesis test

The basic syntax for the `t.test()` function is as follows:

```{r,eval=FALSE}
t.test(x, y = NULL, 
       alternative = c("two.sided", "less", "greater"),
       mu = 0, conf.level = 0.95, ...)
```

* `x`: numeric vector of data values *(our sample)*
* `y`: an *optional* numeric vector of data values *(response)*
* `alternative`: type of hypothesis being performed, 
  - `"two.sided"`: $H_a: \mu \neq \mu_0$
  - `"less"`: $H_a: \mu < \mu_0$
  - `"greater"`: $H_a: \mu > \mu_0$
* `mu`: population mean $\mu$ *(null value)*, can be difference of means if performing a two-sample test
* `conf.level`: confidence level of the interval $100(1-\alpha)$%, where $\alpha$ is the significance level

For additional arguments `...` see `?t.test()`

